---
title: "Final Project Proposal STAT167"
author:
  - "Group Name: Statistically Speaking"
  - "Shreya Mohan, Kalyani Mantirraju, Crystal Arevalo,"
  - "Karen Alvarez, Mason Lam, Eric Yang"
date: "06/02/2025"
format:
  pdf:
    fontsize: 10pt
---

## Libraries
```{r, echo=FALSE}
#install.packages("dunn.test")
library(nycflights13)
library(tidyverse)
library(car)
library(dunn.test)
library(gridExtra)
```

# Project Description:

The primary goal of this research is to explore factors influencing flight delays from New City airports in 2013.

## Problem Statement and Motivation

Understanding factors that contribute to flight delays is critical for informing Federal Aviation Administration (FAA) policies and guiding airlines and airports in improving operational efficiency, enhancing weather preparedness, and reducing delays through controllable factors. By analyzing weather conditions, airline differences, holiday effects, fleet age, and airport specific challenges, this research can provide data-driven insights to optimize air travel and ensure compliance with aviation regulations in heavily congested areas like New York City.

## Research Questions

1.  How do weather conditions affect flight delays?
2.  How do differences between airlines influence flight delays?
3.  Are delays more frequent during major holidays?
4.  Does the age of the plane affect flight delays?
5.  How do environmental factors like humidity, visibility, and wind affect flight delays?
6.  What impact does precipitation have on specific airports and weather-related delays?

# Datasets

## 1. Flights dataset: All flights that departed from NYC in 2013

### Variables:

-   flights ( year, month, day, dep_time, arr_time, sched_dep_time, sched_arr_time, dep_delay, arr_delay, carrier, origin, dest, air_time, distance, time_hour )
    -   year, month, day : date of departure
    -   dep_time, arr_time : actual departure and arrival times in HHMM
    -   sched_dep_time, sched_arr_time : scheduled departure and arrival times in HHMM
    -   dep_delay, arr_delay : departure and arrival delays in minutes
    -   carrier : two letter carrier abbreviation of the carrier
    -   origin, dest : origin and destination
    -   air_time : amount of time spent in air in minutes
    -   distance : distance between airport in miles
    -   time_hour : scheduled date and hour of the flight as POSIXct date

## 2. Airlines dataset: Translation between two letter carrier codes and names

### Variables:

-   airlines ( carrier, name )
    -   carrier : two-letter abbreviation of the airlines
    -   name : full name of the airlines

## 3. Airports dataset: Airport names and locations

```{r}
head(airports)
names(airports)
str(airports)
glimpse(airports)
```

### Variables:

-   airports ( faa, name, lat, lon )
    -   faa : FAA airport code
    -   name : usual name of the airport
    -   lat, lon : location of airport

## 4. Planes dataset: Construction information about each plane

### Variables:

-   planes ( year, type, manufacturer, model, engines, seats, speed, engine )
    -   year : year manufactured
    -   type : type of plane
    -   manufacturer, model : manufacturer and model
    -   engines, seats : number of engines and seats
    -   speed : average cruising speed in mph
    -   engine : type in engine

## 5. Weather dataset: Hourly meterological data for each airport

### Variables:

-   weather ( origin, year, month, day, hour, temp, dewp, humid, wind_dir, wind_speed, wind_gust, precip, pressure, visib, time_hour )
    -   origin : weather station
    -   year, month, day, hour : time of recording
    -   temp, dewp : temperature and dew point in Fahrenheit
    -   humid : relative humidity
    -   wind_dir, wind_speed, wind_gust : wind direction in degrees, wind speed and gust in mph
    -   precip : precipitation in inches
    -   pressure : sea level pressure in millibars
    -   visib : visibility in miles
    -   time_hour : date and hour of the recording as POSIXct date

\newpage
\begin{center}
\section*{Exploratory Data Analysis}
\end{center}

## Planes Dataset EDA

```{r, echo=FALSE}
cat("\nMissing values per column:\n")
print(colSums(is.na(planes)))

cat("\nColumn types and structure:\n")
print(glimpse(planes))

cat("\nFirst few rows:\n")
print(head(planes))

cat("\nSummary statistics:\n")
print(summary(planes))
```

```{r, echo=FALSE}
planes %>%
  count(manufacturer, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(manufacturer, n), y = n)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Top 10 Plane Manufacturers", x = "Manufacturer", y = "Number of Planes")
```

The visualization above shows the top 10 plane manufacturers present in the data-set. Boeing has the largest amount of planes with approximately 1750 planes, and Airbus has the second most with approximately 400 planes.

```{r, echo=FALSE}
ggplot(planes, aes(x = year)) +
  geom_histogram(binwidth = 1, fill = "midnightblue", color = "white") +
  labs(title = "Distribution of Plane Manufacture Years", x = "Year", y = "Count")
```

This histogram shows the distribution of plane manufacture years, with the majority of planes built between the mid-1990s and early 2000s. There is a notable peak around the year 2000, indicating a surge in plane production during that period.

## Airlines Dataset EDA

Dimensions and column names of the airlines dataset

```{r, echo=FALSE}
cat("\nMissing values per column:\n")
print(colSums(is.na(airlines)))

cat("\nColumn types and structure:\n")
print(glimpse(airlines))

cat("\nFirst few rows:\n")
print(head(airlines))

cat("\nSummary statistics:\n")
print(summary(airlines))
```

Viewing all the Unique Airlines:

```{r}
airlines %>%
  arrange(name)
```

```{r}
# Join flights and airline names
flights_airlines <- flights %>%
  left_join(airlines, by = "carrier")

# Average delay metrics
avg_delays <- flights_airlines %>%
  group_by(name) %>%
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    flights = n()
  )

# Plot: Departure Delay
ggplot(avg_delays, aes(x = reorder(name, avg_dep_delay), y = avg_dep_delay)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(
    title = "Average Departure Delay by Airline",
    x = "Airline",
    y = "Avg Departure Delay (min)"
  )
```

We can see that on average, Frontier Airlines has the most departure delay at around 20 min, with ExpressJet roughly around the same 20 minutes. Less than half the Airlines seem to be past the 13 minute delay mark.

```{r}
cancel_rate <- flights_airlines %>%
  mutate(cancelled = is.na(dep_delay)) %>%
  group_by(name) %>%
  summarise(cancel_rate = mean(cancelled), total_flights = n())

ggplot(cancel_rate, aes(x = reorder(name, cancel_rate), y = cancel_rate)) +
  geom_col(fill = "orange") +
  coord_flip() +
  labs(
    title = "Cancelation Rate by Airline",
    x = "Airline",
    y = "Cancelation Rate"
  )
```

As we can see from above, Skywest Airlines Inc has the highest cancellation rate, with Mesa Airlines very closely behind, and a huge drop off at Endeavor Air Inc. \## Planes Dataset EDA

## Flights Dataset EDA:

```{r, echo=FALSE}
cat("\nMissing values per column:\n")
print(colSums(is.na(flights)))

cat("\nColumn types and structure:\n")
print(glimpse(flights))

cat("\nFirst few rows:\n")
print(head(flights))

cat("\nSummary statistics:\n")
print(summary(flights))
```

## Most of our analysis is based on how other variables and datasets affect and compare to the flights dataset. We are seeing how the arrival time, departure delay time, departure time, arrival delay time, and other variables are affected.

#flights that were not canceled \## We will be using these the not_canceled data for the rest of the EDA

```{r}
not_canceled <- filter(flights, !is.na(dep_delay), !is.na(arr_delay))
not_canceled
```

#Basic delay analysis \## Distribution and Proportion of delayed flights that were not canceled

```{r}
#histograms
hist(not_canceled$dep_delay, breaks=100, main = "Departure Delays",xlab = "Minutes")
hist(not_canceled$arr_delay, breaks=100, main ="Arrival Delays", xlab = "Minutes")

#proportions
mean(not_canceled$dep_delay>0, na.rm=TRUE)
mean(not_canceled$arr_delay>0, na.rm=TRUE)
```

#most of the departure delays do not go over 200 minutes and the arrival delays have very few delays past 200 minutes.

#Delay patterns

```{r}
#convert time to hours
not_canceled$dep_hour <-floor(not_canceled$sched_dep_time/100)
not_canceled$arr_hour <-floor(not_canceled$sched_arr_time/100)

#plot
not_canceled |>
  group_by(dep_hour) |>
  summarize(mean_dep_delay = mean(dep_delay, na.rm=TRUE))|>
  ggplot(aes(x=dep_hour, y =mean_dep_delay))+
  geom_line()+
  geom_point()+
  labs(title = "Average Departure Delay by Hour", x="Scheduled Departure", y="Minutes")

```

#We can see that many of the delays happen further in the day and peak at about 18 hours and then it descends from there.

#Delays by Airport

```{r}
not_canceled |>
  group_by(origin)|>
  summarize(avg_dep_delay= mean(dep_delay, na.rm=TRUE), avg_arr_delay= mean(arr_delay, na.rm=TRUE))
```

#EWR has the highest average departure and arrival delay followed by JFK and then LGA

#Ranking airlines by delay

```{r}
not_canceled|>
  group_by(carrier)|>
  summarize(avg_dep_delay = mean(dep_delay, na.rn=TRUE))|>
  arrange(desc(avg_dep_delay))

```

#F9 has the highest average departure delay at 20 hours.

#Check to see if the flights that were delayed made up the time in the air

```{r}
not_canceled|>
  mutate(made_up_time = dep_delay- arr_delay)|>
  ggplot(aes(x=made_up_time))+
  geom_histogram(bins=100)+
  labs(titel="Made up Time in Air", x= "Minutes Saved", y="Count")

```

#We can see that the majority of the flights did not save any minutes on the arrival delay and actually ended up being delayed more. Some flights did in fact save minutes but it was less than 50% of all flights.

#Delays by Month

```{r}
not_canceled|>
  group_by(month)|>
  summarize(mean_dep_delay=mean(dep_delay, na.rm=TRUE))|>
  ggplot(aes(x=month, y=mean_dep_delay))+
  geom_line()+
  geom_point()+
  labs(title = "Monthly Departure Delays", x="Month", y="Time (minutes)")

```

#we see that the majority of flights are delayed from May to mid July, and there is another peak at December. The months with the shorest delays are September and October.

#Does distance affect the amount of delays?

```{r}
ggplot(not_canceled, aes(x=distance, y=dep_delay))+
  geom_point(alpha=0.2)+
  geom_smooth(method = "lm", se=TRUE,color= "Pink")+
  labs(title = "Distance vs Departure Delay", x="Distance", y="Departure Delay")

```

#We can see that there is not much of an effect of Distance on Departure delay.

#Flights traveled the longest by distance

```{r}
longest_Distance <- not_canceled|>
arrange(desc(distance))|>
select(carrier, origin, dest)
longest_Distance
```

#We see that HA is the carrier with the longest flights and they all start at JFK airport and land at HNL.

#arrival delays per carrier

```{r}
carrier_Dest<-not_canceled |>
  group_by(carrier, dest) |>
  summarize(avg_arr_Delay = mean(arr_delay, na.rm= TRUE), .group= "drop")
  ggplot(carrier_Dest, aes(x = reorder(carrier, avg_arr_Delay, median), y= avg_arr_Delay))+
  geom_boxplot()+
  coord_flip()+
  labs(title = "Average Arrival Delay by Carrier",x = "Carrier", y= "Average Arrival Delay (minutes)")+
  theme_minimal()
```

## Weather Dataset EDA

```{r, echo=FALSE}
cat("\nMissing values per column:\n")
print(colSums(is.na(weather)))

cat("\nColumn types and structure:\n")
print(glimpse(weather))

cat("\nFirst few rows:\n")
print(head(weather))

cat("\nSummary statistics:\n")
print(summary(weather))
```

```{r}
# Related Questions from our Proposal:
# --- How do weather conditions affect flight delays? 
# --- How do environmental factors like humidity, visibility, and wind affect flight delays? 
# --- What impact does precipitation have on specific airports and weather-related delays? 


# Average Temperature by Month
# --- We can see that the average temperature ranges from about 70-80º in the summer, and 35-40º in the winter. 
# --- When answering our research questions, we can see if there is a correlation between summer/winter weather and flight delays.

monthlyavgtemp <- weather %>%
  group_by(month) %>%
  summarise(monthlyavgtemp = mean(temp, na.rm = TRUE))

ggplot(data = monthlyavgtemp,
       aes(x = factor(month), y = monthlyavgtemp)) +
  geom_col(color = "skyblue", fill = "lightskyblue1") + 
  labs(title = "Average Temperature by Month",
       x = "Month",
       y = "Average Temperature (Fº)") +
  theme_minimal()
  

# Average Visibility by Month
# --- The average visibility does not greatly vary by month looking at the average value. However, we can see that there is slightly less visiblity in winter months.
# --- Looking at the boxplots, we can see that there are a lot of outliers. Removing these outliers and focusing on visib < 10 shows us a better distribution of visiblity.
# --- When answering our research questions, we can compare the average visibility during flight delays vs average visibility without flight       delays to further explore the role of visibility in flight delays. 

monthlyavgvisib <- weather %>%
  group_by(month) %>%
  summarise(monthlyavgvisib = mean(visib, na.rm = TRUE))

ggplot(data = monthlyavgvisib) +
    geom_col(aes(x = factor(month), y = monthlyavgvisib),
                   color = "orchid4", fill = "plum") + 
  labs(title = "Average Visibility by Month",
       x = "Month",
       y = "Average Visibility") +
  theme_minimal()

ggplot(weather, aes(x = factor(month), y = visib)) +
  geom_boxplot(fill = "plum", color = "orchid4") +
  labs(title = "Distribution of Visibility by Month",
       x = "Month",
       y = "Visibility") +
  coord_flip() +
  theme_minimal()

ggplot(filter(weather, visib < 10), aes(x = factor(month), y = visib)) +
  geom_boxplot(fill = "plum", color = "orchid4") +
  labs(title = "Distribution of Visibility < 10 by Month",
       x = "Month",
       y = "Visibility") +
  coord_flip() +
  theme_minimal()


# Distribution of Humidity by Month
# --- The distribution of humidity varies by month, but there does not seem to be significant differences.
# --- We can further explore the role of humidity by comparing it to other weather variables and flight delays.

ggplot(weather, aes(x = factor(month), y = humid)) +
  geom_boxplot(fill = "darkseagreen2", color = "darkgreen") +
  labs(title = "Distribution of Humidity by Month (With Outliers)",
       x = "Month",
       y = "Humidity") +
  coord_flip() +
  theme_minimal()


# Precipitation by Month
# --- The average precipitation for each month varies greatly. We can see that spring months have the greatest average precipitation.
# --- When answering our research question, we can see if greater precipitation correlates to flight delays. 

monthlyprecip <- weather %>%
  group_by(month) %>%
  summarise(avgmonthlyprecip = mean(precip, na.rm = TRUE))

ggplot(monthlyprecip, aes(x = month, y = avgmonthlyprecip)) +
  geom_line(color = "palevioletred2") +
  geom_point(color = "palevioletred4") +
  labs(title = "Average Monthly Precipitation",
       x = "Month",
       y = "Average Precipitation") +
  scale_x_continuous(breaks = 1:12) +
  theme_minimal()


# Correlation Between Variables
# --- We can explore the correlation between different weather variables and see how they may work together to impact flight delays.

cor(weather$precip, weather$visib, use = "complete.obs")
cor(weather$humid, weather$visib, use = "complete.obs")


```

\newpage

# Analysis Approach Plan

**Assumptions:** All variables are independent

The process of analysis will involve data cleaning after forming our question, basic exploration of the data, comparison of certain datasets with other datasets, visualization of the data, and an interpretation of the data/results. Cleaning of the data will deal with tasks like handling empty cells/columns and NA values. When it comes to exploratory data analysis, we plan on using tools such as histograms and boxplots to gain an understanding of the data and identify patterns and relationships. The statistical analysis that we plan on performing with the data will most likely involve making comparisons between groups to compare airlines, times, and other metrics to make our overall claim. For example, we might be comparing trends in time performance by weeks or month between different airlines to gain a better understanding of how differences in airlines affect delays. In terms of data visualization, we will most likely be using line graphs for trends over time when it comes to comparing flight time under different variables and heatmaps/scatterplots for flight delays to help communicate our findings. Finally, interpretation of the data will involve us answering the proposed question by summarizing our statistics/findings as well as through the presentation of graphical evidence.

# Analysis:

### Question 4: Does the age of the plane affect flight delays?

Within this main question we will perform hypothesis tests to answer the two following sub-questions:

#### 1.  Do older planes experience more delays compared to newer ones? 

$H_0$: There is no difference in the distribution of arrival delays across the different plane age groups.

$H_1$: At least one plane age group has a different distribution of arrival delays compared to the others.

```{r}
# Join flights with planes to get plane manufacture year
planes_fixed <- planes %>%
  rename(plane_year = year)

flights_planes <- flights %>%
  inner_join(planes %>% rename(plane_year = year), by = "tailnum") %>%
  filter(!is.na(plane_year), !is.na(arr_delay)) %>%
  mutate(
    plane_age = 2013 - plane_year,
    age_group = case_when(
      plane_age < 10 ~ "0-9 yrs",
      plane_age < 20 ~ "10-19 yrs",
      plane_age < 30 ~ "20-29 yrs",
      TRUE ~ "30+ yrs"
    )
  )

head(flights_planes)
```

```{r, echo=FALSE}
ggplot(flights_planes, aes(x = age_group, y = arr_delay)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Arrival Delay by Plane Age Group",
    x = "Plane Age Group",
    y = "Arrival Delay (minutes)"
  )
```
The boxplot shows that arrival delays are fairly similar across all plane age groups, with comparable medians and interquartile ranges. While each group has extreme outliers, there is no clear visual trend suggesting that older planes experience more delays than newer ones. We will run a hypothesis test to further our findings. 

```{r}
flights_planes$age_group <- as.factor(flights_planes$age_group)

summary_stats <- flights_planes %>%
  group_by(age_group) %>%
  summarise(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    count = n()
  )
print(summary_stats)

# Normality test with sample per group
set.seed(123)
normality_test <- flights_planes %>%
  group_by(age_group) %>%
  summarise(
    sample_delays = list(sample(arr_delay[!is.na(arr_delay)], min(5000, n()), replace = FALSE)),
    shapiro_p = shapiro.test(unlist(sample_delays))$p.value
  ) %>%
  select(-sample_delays)
print(normality_test)

# Levene's Test for homogeneity of variances
leveneTest(arr_delay ~ age_group, data = flights_planes)

# Since homogeneity is violated, run Kruskal-Wallis test (non-parametric)
kruskal.test(arr_delay ~ age_group, data = flights_planes)

# post-hoc test for Kruskal-Wallis (Dunn test) if significant
dunn.test(flights_planes$arr_delay, flights_planes$age_group, method = "bonferroni")

```

The Kruskal-Wallis test revealed a highly significant difference in arrival delays across plane age groups (p \< 2.2e-16), indicating that at least one group’s delay distribution differs from the others. Post-hoc pairwise comparisons using Dunn’s test with Bonferroni correction showed that planes aged 20–29 years experience significantly different delay patterns compared to both the 0–9 and 10–19 year groups (adjusted p-values \< 0.001). Additionally, planes aged 30+ years differ significantly from the 20–29 year group (adjusted p = 0.0002), but do not differ significantly from the younger 0–9 or 10–19 year groups. The difference between the 10–19 and 0–9 year groups was borderline significant (adjusted p = 0.0495). In summary, planes aged 20–29 years tend to have notably different arrival delays compared to most other age groups, highlighting a possible link between this age range and on-time performance issues.

#### 2.  Are there specific plane models or manufactures associated with better on-time performance? 

$H_0$: There is no difference in the distribution of arrival delays across different plane models or manufacturers.

$H_1$: At least one plane model or manufacturer has a different distribution of arrival delays compared to the others.

```{r}
flights_manufacturer <- flights %>%
  inner_join(planes, by = "tailnum") %>%
  filter(!is.na(manufacturer), !is.na(arr_delay)) %>%
  group_by(manufacturer) %>%
  filter(n() > 50)
```

```{r, echo=FALSE}
ggplot(flights_manufacturer, aes(x = reorder(manufacturer, arr_delay, FUN = median), y = arr_delay, fill = manufacturer)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  coord_flip() +   # horizontal for readability
  labs(
    title = "Arrival Delay Distribution by Plane Manufacturer",
    x = "Manufacturer",
    y = "Arrival Delay (minutes)"
  )
```

```{r}
# filter manufacturers with more than 50 flights
flights_manufacturer <- flights %>%
  inner_join(planes, by = "tailnum") %>%
  filter(!is.na(manufacturer), !is.na(arr_delay)) %>%
  group_by(manufacturer) %>%
  filter(n() > 50) %>%
  ungroup()

flights_manufacturer$manufacturer <- as.factor(flights_manufacturer$manufacturer)

# Normality test per manufacturer group
set.seed(123)
normality_test <- flights_manufacturer %>%
  group_by(manufacturer) %>%
  summarise(
    sample_delays = list(sample(arr_delay, min(5000, n()), replace = FALSE)),
    shapiro_p = shapiro.test(unlist(sample_delays))$p.value
  ) %>%
  select(-sample_delays)
print(normality_test)

# Levene's Test for homogeneity of variance
leveneTest(arr_delay ~ manufacturer, data = flights_manufacturer)

# Kruskal-Wallis test 
kruskal_result <- kruskal.test(arr_delay ~ manufacturer, data = flights_manufacturer)
print(kruskal_result)

if (kruskal_result$p.value < 0.05) {
  dunn_result <- dunn.test(flights_manufacturer$arr_delay, flights_manufacturer$manufacturer, method = "bonferroni")
  
  sig_comparisons <- data.frame(
    comparison = dunn_result$comparisons,
    p_value = dunn_result$P.adjusted
  ) %>% 
    filter(p_value < 0.05) %>%
    arrange(p_value) %>%
    slice_head(n = 5) %>%  # Show only top 5
    mutate(p_value = signif(p_value, 4))  # Optional: round p-values
  
  cat("Kruskal-Wallis p-value:", signif(kruskal_result$p.value, 4), "\n")
  cat("Significant differences found. Top 5 manufacturer pairs with different arrival delays:\n\n")
  print(sig_comparisons, row.names = FALSE)
  
} else {
  cat("Kruskal-Wallis p-value:", signif(kruskal_result$p.value, 4), "\n")
  cat("No significant differences found among manufacturers.\n")
}
```

The Kruskal-Wallis test indicates a highly significant difference in arrival delays across plane manufacturers (χ² = 3770.9, df = 21, p < 2.2e-16). This strongly suggests that not all manufacturers have the same on-time performance. The post-hoc Dunn's test with Bonferroni correction highlights specific pairwise differences. Notably, comparisons involving Embraer show consistently significant differences with Airbus Industrie, Boeing, Bombardier Inc, and McDonnell Douglas, indicating that Embraer aircraft tend to have distinct arrival delay patterns compared to these manufacturers. Additionally, Airbus and McDonnell Douglas pairs also appear among the most significant differences. These results indicate that certain manufacturers, especially Embraer, are associated with notably different arrival delay profiles, reflecting differences in punctuality or operational factors across manufacturers.

\newpage

# Conclusion

# Member Contributions

# Alternative Strategies & Back Up Plan:

As a backup idea, we are planning on seeing if there is any correlation between the amount of delays present in the different airports. Our data deals with the airports EWR, JFK, and LGA which are all different airports within New York City. Our first question is to figure out if the JFK airport has a different amount of delays compared to LGA or EWR if there is a higher amount of precipitation in the JFK area. Although all the airports are in New York, within the different areas of the city, there can be different amounts of precipitation and rainfall that occur. Our second question is to decide whether the different airports have different models of planes and if the difference affects the amounts of delays. For example if a plane is older or a different configuration, does that lead to more delays due to cleaning or maintenance? And lastly, our third question is whether the three different airports have different airlines coming in and out and if these differing airlines affect the amount of delays present on a given day. For example, if Delta services one airport and not another, does that increase or decrease the amount of total delays for an airport. These questions can be further investigated if our first set of questions are not approved or if we need more content to explore within our project. These sets of backup questions will further explore the flight data we have.
