---
title: "Final Project Proposal STAT167"
author: 
  - "Group Name: Statistically Speaking"
  - "Shreya Mohan, Kalyani Mantirraju, Crystal Arevalo, Karen Alvarez, Mason Lam"
date: "04/27/2025"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-line-numbers: true
  pdf:
    code-line-numbers: true
    fontsize: 10pt
---


# Installation & Packages

```{r}
library(nycflights13)
library(tidyverse)
```

# Introduction

The primary goal of this research is to explore factors influencing flight delays from New City airports in 2013.

# Problem Statement and Motivation
Understanding factors that contribute to flight delays is critical for informing Federal Aviation Administration (FAA) policies and guiding airlines and airports in improving operational efficiency, enhancing weather preparedness, and reducing delays through controllable factors. By analyzing weather conditions, airline differences, holiday effects, fleet age, and airport specific challenges, this research can provide data-driven insights to optimize air travel and ensure compliance with aviation regulations in heavily congested areas like New York City. 

# Main Research Question
What are the key correlations between flight delays from NYC airports?

## Sub-questions:

The following questions will guide the analysis:

1.  How do weather conditions affect flight delays?
    a.  Are specific weather variables (e.g., precipitation, wind speed, humidity) correlated with departure and arrival delays?
2.  How do differences between airlines influence flight delays?
    a.  Do certain airlines experience more delays than other, if so, what operational or fleet-related factors contribute to these differences?
    b.  How do metrics like cancelation rates and plane speed vary across airlines, and what impact do these metrics have on delays?
3.  Are delays more frequent during major holidays?
    a.  Are there differences during peak travel periods (e.g., Thanksgiving, Christsmas, New Year’s Day)
4.  Does the age of the plane affect flight delays?
    a.  Do older planes experience more delays compared to newer ones?
    b.  Are there specific plane models or manufactures associated with better on-time performance?
5.  How do environmental factors like humidity, visibility, and wind affect flight delays?
    a.  Are these effects observed across all airports?
6.  What impact does precipitation have on specific airports and weather-related delays?
    a.  Do airports in regions with higher average precipitation experience more delays?

# Datasets

## 1. Flights dataset: All flights that departed from NYC in 2013
```{r}
head(flights)
dim(flights)
names(flights)
str(flights)
glimpse(flights)
```

### Variables:

-   flights ( year, month, day, dep_time, arr_time, sched_dep_time, sched_arr_time, dep_delay, arr_delay, carrier, origin, dest, air_time, distance, time_hour )
    -   year, month, day : date of departure
    -   dep_time, arr_time : actual departure and arrival times in HHMM
    -   sched_dep_time, sched_arr_time : scheduled departure and arrival times in HHMM
    -   dep_delay, arr_delay : departure and arrival delays in minutes
    -   carrier : two letter carrier abbreviation of the carrier
    -   origin, dest : origin and destination
    -   air_time : amount of time spent in air in minutes
    -   distance : distance between airport in miles
    -   time_hour : scheduled date and hour of the flight as POSIXct date

## 2. Airlines dataset: Translation between two letter carrier codes and names
```{r}
head(airlines)
dim(airlines)
names(airlines)
str(airlines)
glimpse(airlines)
```

### Variables:

-   airlines ( carrier, name )
    -   carrier : two-letter abbreviation of the airlines
    -   name : full name of the airlines

## 3. Airports dataset: Airport names and locations
```{r}
head(airports)
dim(airports)
names(airports)
str(airports)
glimpse(airports)
```

### Variables:

-   airports ( faa, name, lat, lon )
    -   faa : FAA airport code
    -   name : usual name of the airport
    -   lat, lon : location of airport
    
## 4. Planes dataset: Construction information about each plane
```{r}
head(planes)
dim(planes)
names(planes)
str(planes)
glimpse(planes)
```

### Variables:

-   planes ( year, type, manufacturer, model, engines, seats, speed, engine )
    -   year : year manufactured
    -   type : type of plane
    -   manufacturer, model : manufacturer and model
    -   engines, seats : number of engines and seats
    -   speed : average cruising speed in mph
    -   engine : type in engine
    
## 5. Weather dataset: Hourly meterological data for each airport
```{r}
head(weather)
dim(weather)
names(weather)
str(weather)
glimpse(weather)
```

### Variables:

-   weather ( origin, year, month, day, hour, temp, dewp, humid, wind_dir, wind_speed, wind_gust, precip, pressure, visib, time_hour )
    -   origin : weather station
    -   year, month, day, hour : time of recording
    -   temp, dewp : temperature and dew point in Fahrenheit
    -   humid : relative humidity
    -   wind_dir, wind_speed, wind_gust : wind direction in degrees, wind speed and gust in mph
    -   precip : precipitation in inches
    -   pressure : sea level pressure in millibars
    -   visib : visibility in miles
    -   time_hour : date and hour of the recording as POSIXct date

We can join the data tables by combining them through similar attributes, such as combining flights : time_hour with weather : time_hour.


## EDA:
## Summary Statistics & Check for Missing Values:
```{r}
summary(flights)
colSums(is.na(flights))

summary(airlines)
colSums(is.na(airlines))

summary(airports)
colSums(is.na(airports))

summary(planes)
colSums(is.na(planes))

summary(weather)
colSums(is.na(weather))
```



# Analysis Approach Plan:

**Assumptions:** All variables are independent

The process of analysis will involve data cleaning after forming our question, basic exploration of the data, comparison of certain datasets with other datasets, visualization of the data, and an interpretation of the data/results. Cleaning of the data will deal with tasks like handling empty cells/columns and NA values. When it comes to exploratory data analysis, we plan on using tools such as histograms and boxplots to gain an understanding of the data and identify patterns and relationships. The statistical analysis that we plan on performing with the data will most likely involve making comparisons between groups to compare airlines, times, and other metrics to make our overall claim. For example, we might be comparing trends in time performance by weeks or month between different airlines to gain a better understanding of how differences in airlines affect delays. In terms of data visualization, we will most likely be using line graphs for trends over time when it comes to comparing flight time under different variables and heatmaps/scatterplots for flight delays to help communicate our findings. Finally, interpretation of the data will involve us answering the proposed question by summarizing our statistics/findings as well as through the presentation of graphical evidence.

# Alternative Strategies & Back Up Plan:

As a backup idea, we are planning on seeing if there is any correlation between the amount of delays present in the different airports. Our data deals with the airports EWR, JFK, and LGA which are all different airports within New York City. Our first question is to figure out if the JFK airport has a different amount of delays compared to LGA or EWR if there is a higher amount of precipitation in the JFK area. Although all the airports are in New York, within the different areas of the city, there can be different amounts of precipitation and rainfall that occur. Our second question is to decide whether the different airports have different models of planes and if the difference affects the amounts of delays. For example if a plane is older or a different configuration, does that lead to more delays due to cleaning or maintenance? And lastly, our third question is whether the three different airports have different airlines coming in and out and if these differing airlines affect the amount of delays present on a given day. For example, if Delta services one airport and not another, does that increase or decrease the amount of total delays for an airport. These questions can be further investigated if our first set of questions are not approved or if we need more content to explore within our project. These sets of backup questions will further explore the flight data we have.
